---
title: "Transcriptome Demo"
author: "Mason"
date: "`r Sys.Date()`"
output: pdf_document
---

# Load required packages (you might have to figure out how to install some of these first...)
```{r message=FALSE}
library(ballgown)
library(RColorBrewer)
library(genefilter)
library(dplyr)
library(devtools)
```

# What is this code doing? it creating a data frame called pheno_data that stores sample metadata (also known as phenotype data), which will be used to associate each sample with its biological condition (planktonic or biofilm). This is important for differential expression analysis in transcriptomics.
```{r}
pheno_data<-data.frame(ids = c("plank01", "plank02", "biofilm01", "biofilm02"),
                       stage = c("planktonic", "planktonic", "biofilm", "biofilm"))
```

# create Ballgown object and check transcript number
```{r message=FALSE}
load("bg.RData")
bg
```

# <what is this code doing?> This is a common filtering step before running differential expression, helping to eliminate uninformative or low-variance transcripts that could bias results.


```{r}
bg_filt = subset(bg,"rowVars(texpr(bg)) >1",genomesubset=TRUE)
bg_filt
```

# create a table of transcripts
```{r}
results_transcripts<- stattest(bg_filt, feature = "transcript", covariate = "stage",
getFC = TRUE, meas = "FPKM")
results_transcripts<-data.frame(geneNames=geneNames(bg_filt),
transcriptNames=transcriptNames(bg_filt), results_transcripts)
```

# choose a transcript to examine more closely (this is a demo, you need to choose another)
```{r}
results_transcripts[results_transcripts$transcriptNames == "gene-PA0002
", ]
```
## what information are you given about this transcript? dentifying which transcripts show statistically significant changes in expression between biofilm and planktonic conditions. The output from dim(sigdiff) tells you how many such transcripts were found.

# <insert brief explanation of this code>
```{r}
sigdiff <- results_transcripts %>% filter(pval<0.05)
dim(sigdiff)
```

# organize the table <by what metrics is the table being organized?> organizing the sigdiff table by p-value first (smallest to largest), and then by absolute fold change (largest to smallest),
```{r}
o = order(sigdiff[,"pval"], -abs(sigdiff[,"fc"]), decreasing=FALSE)
output = sigdiff[o,c("geneNames","transcriptNames", "id","fc","pval","qval")]
write.table(output, file="SigDiff.txt", sep="\t", row.names=FALSE, quote=FALSE)
head(output)
```

# load gene names
```{r}
bg_table = texpr(bg_filt, 'all')
bg_gene_names = unique(bg_table[, 9:10])
```

# pull out gene expression data and visualize
```{r}
gene_expression = as.data.frame(gexpr(bg_filt))
head(gene_expression)
```

# <what is this code doing? hint:compare the above output of head(gene_expression) to this output> This code is renaming the columns of the gene_expression data frame to match the sample IDs and then displaying the first few rows and dimensions of the modified data.
```{r}
colnames(gene_expression) <- c("plank01", "plank02", "biofilm01", "biofilm02")
head(gene_expression)
dim(gene_expression)
```

# load the transcript to gene table and determine the number of transcripts and unique genes
```{r}
transcript_gene_table = indexes(bg)$t2g
head(transcript_gene_table)
length(row.names(transcript_gene_table))
length(unique(transcript_gene_table[,"g_id"]))
```

# plot the number of transcripts per gene
```{r}
counts=table(transcript_gene_table[,"g_id"])
c_one = length(which(counts == 1))
c_more_than_one = length(which(counts > 1))
c_max = max(counts)
hist(counts, breaks=50, col="bisque4", xlab="Transcripts per gene",
main="Distribution of transcript count per gene")
legend_text = c(paste("Genes with one transcript =", c_one),
paste("Genes with more than one transcript =", c_more_than_one),
paste("Max transcripts for single gene = ", c_max))
legend("topright", legend_text, lty=NULL)
```

## <insert your interpretation of the above graph> Most of the genes in this transciptome dataset are simple with the 4128 gene producing only a single transcript. This may suggest limited alternative splicing or transcriptvariation for the genes in P. aerugnosa. Smaller transcript like the 587 genes indicates transcriptional complexity or variation. The 74 distinct transcripts has the highest complexity. 

# create a plot of how similar the two replicates are for one another. We have two data sets...how can you modify this code in another chunk to create a plot of the other set? By replacing the plank01 and plank02 with biofilm01 and biofilm02.
```{r}
x = gene_expression[,"plank01"]
y = gene_expression[,"plank02"]
min_nonzero=1
plot(x=log2(x+min_nonzero), y=log2(y+min_nonzero), pch=16, col="purple", cex=0.25,
xlab="FPKM (Planktonic, Replicate 1)", ylab="FPKM (Planktonic, Replicate 2)",
main="Comparison of expression values for a pair of replicates")
abline(a=0,b=1, col = "hotpink")
rs=cor(x,y)^2
legend("topleft", paste("R squared = ", round(rs, digits=3), sep=""), lwd=1, col="black")
```

## What does it mean if the two data sets are similar? It would mean that the biological replicates are consistent and reproducible this means identical gene expression patterns. This shows consistency, minimal error, and good expresion. 

# create plot of differential gene expression between the conditions
```{r}
results_genes = stattest(bg_filt, feature="gene", covariate="stage", getFC=TRUE, meas="FPKM")
results_genes = merge(results_genes,bg_gene_names,by.x=c("id"),by.y=c("gene_id"))
sig=which(results_genes$pval<0.05)
results_genes[,"de"] = log2(results_genes[,"fc"])
hist(results_genes[sig,"de"], breaks=50, col="seagreen",
xlab="log2(Fold change) Planktonic vs Biofilm",
main="Distribution of differential expression values")
abline(v=-2, col="black", lwd=2, lty=2)
abline(v=2, col="black", lwd=2, lty=2)
legend("topleft", "Fold-change > 4", lwd=2, lty=2)
```

## interpret the above figure: This shows frequency over log2(Fold change) Planlktinic vs Biofilm. There is a peak around the negative numbers just before 0 showing a decrease abundance in biofilm condition with an increase in features and the peak drops at and around 0 because both decrease abundance and decrease in features of and this slope grows but then decreases continues decreasing going into the positive numbers because of leaveling out in data after log 2 grows. 

# Plot total gene expression highlighting differentially expressed genes
```{r}
gene_expression[,"plank"]=apply(gene_expression[,c(1:2)], 1, mean)
gene_expression[,"biofilm"]=apply(gene_expression[,c(3:4)], 1, mean)
x=log2(gene_expression[,"plank"]+min_nonzero)
y=log2(gene_expression[,"biofilm"]+min_nonzero)
plot(x=x, y=y, pch=16, cex=0.25, xlab="Planktonic FPKM (log2)", ylab="Biofilm FPKM (log2)",
main="Planktonic vs Biofilm FPKMs")
abline(a=0, b=1)
xsig=x[sig]
ysig=y[sig]
points(x=xsig, y=ysig, col="magenta", pch=16, cex=0.5)
legend("topleft", "Significant", col="magenta", pch=16)
```

# make a table of FPKM values
```{r}
fpkm = texpr(bg_filt,meas="FPKM")
```

# choose a gene to determine individual expression (pick a different number than I did) index 3
```{r}
ballgown::transcriptNames(bg_filt)[3]
ballgown::geneNames(bg_filt)[3]
```
# transform to log2
```{r}
transformed_fpkm <- log2(fpkm[3, ] + 1)
```

# make sure values are properly coded as numbers
```{r}
numeric_stages <- as.numeric(factor(pheno_data$stage))

jittered_stages <- jitter(numeric_stages)
```

# plot expression of individual gene
```{r}
boxplot(transformed_fpkm ~ pheno_data$stage,
        main=paste(ballgown::geneNames(bg_filt)[3], ' : ', ballgown::transcriptNames(bg_filt)[3]),
        xlab="Stage",
        ylab="log2(FPKM+1)",
        col=c("lightblue", "salmon"),
        border="darkblue")

points(transformed_fpkm ~ jittered_stages, 
       pch=21, col="blue", bg="lightblue", cex=1.2)
```

## interpret the above figure: This figure shows log2(FPKM+1) vs biofilm and planktonic. It shows biofilm values for log2 from around 8.0-8.4 which represents a slightly lower expression and planktonic values around 8.7-9.0 which shows higher expression of the (3) gene. This all means that the planktonic state is more 
